 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-07-09 paper.Rmd
update bibliography and add info on neighbors

### Tours explore the subspace $d$-dimension projections

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ the tour is visualised as
a animated scatterplot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flipbook" 
animation. Like with a flipbook, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in `liminal`
allows a user to play and pause the tour animation (via brushing),
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points (like clusters).

These interactions are enabled by treating the basis generation as a reactive
stream. Instead of realising the entire sequence, which limits the animation
to have a discrete number of frames, new bases and their intermediate frames
are generated dynamically via pushing the current projection to the 
visualisation interface. The interface listens to events like pressing a button
or mouse-dragging and reacts by pausing the stream. This process allows the
user to manipulate the tour in real time rather than having to fix the number
of bases ahead of time.

### Linking views via brushes

- linking multiple views

@@

The next section steps through three use cases of our approach using simulated
and real data. In the first case study, we step through data that lies on tree
structure (similar to trajectories in single cell transcriptomics) to that
highlights an iterative approach to modifying t-SNE views based on what we
learn in the high-dimensional space. We show how through simple linked
brushing, we can see when the assumptions of continuity of the embedding map
are broken. In the second case study, we explore a well-clustered single cell
transcriptomics data of mice retina... Finally, in the last case study we
explore a subset of the QuickDraw dataset, we use `liminal` to identify
instances of images from different labels that appear similar.

@@

In the final case study, we explore a subset of the Google QuickDraw dataset 
[@quickdraw]. The data are sketches drawn by people of common concepts
like 'airplane' or 'shoe', and corresponding metadata about the humans who
created the sketches, such as country, the word category, whether the sketch
was recognised by Google's neural network, and the number of strokes taken.
We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the 
`quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into 
a 784 dimension image vector. We sampled 500 images from each of the following
categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
and "whale".

>>>>

### Tours explore the subspace of $d$-dimensional projections

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ the tour is visualised as
a animated scatterplot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flipbook" 
animation. Like with a flipbook, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in `liminal`
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

These interactions are enabled by treating the basis generation as a reactive
stream. Instead of realising the entire sequence, which limits the animation
to have a discrete number of frames, new bases and their intermediate frames
are generated dynamically via pushing the current projection to the 
visualisation interface. The interface listens to events like pressing a button
or mouse-dragging and reacts by pausing the stream. This process allows the
user to manipulate the tour in real time rather than having to fix the number
of bases ahead of time. Additionally, once the user has identified an 
interesting projection or is done with the tour, the interface will return the 
current basis for use downstream.

### Linking and highlighting views via interactions

The embedding and tour views are linked together via rectangular brushes;
when a brush is active, points will be highlighted in the adjacent view. 
Because the tour is dynamic, brush events that become active will pause the
animation, so that a user can interrogate the current view. By default,
brushing on the embedding view will produce a one-to-one linking with the tour
view (the special case of $k=1$ nearest neighbors brush). It is possible
to shift to one-to-many linking by increasing $k$ via the 'controls'
panel of the interface.  If desired, the distance
metric used to estimate nearest neighbours can by chosen by the user. 
The user may provide a reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix)
independently of the tour or embedding views. The estimation of neighbours
is done with the `FNN` package for datasets with less than 4096 observations
and `RcppAnnoy` for larger datasets or if a user selects a non-euclidean
distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
with the interface, the brush regions are returned for downstream analysis.

@@

The next section steps through three use cases of our approach using simulated
and real data. In the first case study, we step through three simulated datasets
to show the `liminal` interface. The first simulation contains 3 clusters
sampled from a 5-d gaussian distribution that are well embedded by t-SNE.
The second is another example of spherical clusters but this time, they are
tightly clustered and are distorted by t-SNE, which is immediately apparent in 
the tour view. In the final simulation we
look at a tree structure (similar to trajectories in single cell 
transcriptomics) that
highlights an iterative approach to modifying t-SNE views based on what we
learn in the high-dimensional space. We show how through simple linked
brushing, we can see when the assumptions of continuity of the embedding map
are broken.

In the second case study, we explore a well-clustered single cell
transcriptomics data of mice retina...

Finally, in the last case study we
explore a subset of the QuickDraw dataset, we use `liminal` to identify
instances of images from different labels that appear similar.

@@

In the final case study, we explore a subset of the Google QuickDraw dataset 
[@quickdraw-paper]. The data are sketches drawn by people of common concepts
like 'airplane' or 'shoe', and corresponding metadata about the humans who
created the sketches, such as country, the word category, whether the sketch
was recognised by Google's neural network, and the number of strokes taken.
We downloaded the metadata and normalised 28 by 28 pixel bitmaps via the 
`quickdraw` package in R; unravelling the bitmaps unravelled the bitmaps into 
a 784 dimension image vector. We sampled 500 images from each of the following
categories: "broccoli", "flying saucer", "ice cream", "octopus", "pineapple",
and "whale".
